Time stretching of the GeV emission of GRBs: Fermi LAT data vs geometrical model

### 2014-11-03 Paper/gammaRays.tex
Abstract revised

\usepackage{commath}
\usepackage{graphicx}
\usepackage{fullpage}
\usepackage{subcaption}
\usepackage{units}
\usepackage{hyperref}
\usepackage{siunitx}
\usepackage{listings}
\usepackage{multicol}
\usepackage{bm}
\usepackage{appendix}
\usepackage{textcomp}
\usepackage{gensymb}

@@

\title{
	Time Stretching of the GeV Emission of GRBs: Fermi LAT data vs geometrical model \\
	{\large \url{https://github.com/maxitg/GammaRays}}
}

\author{
	Maxim S. Piskunov$^{1}$\thanks{\href{mailto:maxit@ms2.inr.ac.ru}{maxit@ms2.inr.ac.ru}} and
	Grigory I. Rubtsov$^{1}$\thanks{\href{mailto:grisha@ms2.inr.ac.ru}{grisha@ms2.inr.ac.ru}} \\
	$^{1}$Institute for Nuclear Research RAS, 117312, Moscow, Russia\\
}

@@

\begin{abstract}
	Numerous observations confirm that the high energy $(> \unit[100]{MeV})$ emission of gamma ray bursts is delayed with respect to the low energy emission.
	However, the difference of light curves in various high energy bands has not been studied properly.

	In this paper we consider all the bursts observed by Fermi-LAT since 2008 August 4 to 2011 August 1, for which at least $10$ events with energies $\unit[1]{GeV}$ or higher were observed.
	There are $4$ of them: GRB 080916C, GRB 090510, GRB 090902B, and GRB 090926A.
	We study their light curves in two bands, $(\unit[100]{MeV}, \unit[1]{GeV})$ and $(\unit[1]{GeV}, \unit[300]{GeV})$.

	The Kolmogorov-Smirnov test is used to check whether the light curves for these two bands are the same.
	No significant difference was found for GRB 080916C and GRB 090510.
	However, we observed with statistical significance of $3.3 \sigma$, that the higher energy light curve of GRB 090926A is stretched with respect to the lower-energy one, and with statistical significance of $2.3 \sigma$, that the lower energy light curve of GRB 090902B is stretched with respect to the higher-energy one.

	We suggest a simple geometrical model to explain this result.
	The main assumption is the jet opening angle dependence on radiation energy -- the most energetic photons are emitted near the axis of the jet.
	We also assume that all bursts are the same in their rest frames (that is their light curves differ only because of different redshifts and different view directions).
	To test this model, we compute the total energy of the burst, and confirm that it is below the constraint.
	We also compute the fraction of observable bursts in $(\unit[100]{MeV}, \unit[1]{GeV})$ band, which can also be observed in higher energies.
	This fraction matches the observations.
	We predict the distribution of observable stretching factors, which may be tested in the future when more observational data become available.
	Finally, we propose a way to estimate observer's off-axis angles (view directions) based on stretching factors and fractions of high energy photons.

	All our code is open source, and is available freely on GitHub for you to download and play with.
\end{abstract}

@@

	Gamma Ray Bursts (GRBs) are among the most energetic events in the Universe, therefore they might provide new knowledge for particle physics.
	Modern observatories, specifically Fermi \cite{Ackermann:2012kna} and Swift \cite{Gehrels:2004gu} made possible to study these explosions extensively \cite{Vianello:2013ela,Gehrels:2013xd}.
	These and previous studies led to several interesting results.
	For example, the total energy emitted in gamma-rays during a burst was found to be similar among different bursts within an order of magnitude \cite{Bloom:2003wy}.
	This suggests that most of the bursts must have similar energetics in their rest frames.
	Temporal variations of spectra were also studied. The spectral lags were found between different low energy bands \cite{Yi:2005ht}.
	The very high energy radiation was discovered to be extended relative to x-ray emission \cite{Castignani:2014gaa,Lange:2013uh,Vianello:2013ela}.
	We decided to elaborate on that result, and, using the data from Large Area Telescope (LAT) of Fermi, explore the spectral variations between different high energy bands, specifically $(\unit[100]{MeV}, \unit[1]{GeV})$ and $(\unit[1]{GeV}, \unit[300]{GeV})$ (we call them low and high energy bands throughout the paper).
	In particular, we use the Kolmogorov-Smirnov test to compute the time stretching of radiation in one of these bands compared to the other.

@@

\bibliographystyle{hep}
\bibliography{gammaRays.bib}

>>>>

\usepackage{commath}
\usepackage{graphicx}
\usepackage{fullpage}
\usepackage{subcaption}
\usepackage{units}
\usepackage{hyperref}
\usepackage{siunitx}
\usepackage{listings}
\usepackage{multicol}
\usepackage{bm}
\usepackage{appendix}
\usepackage{textcomp}
\usepackage{gensymb}
\usepackage{booktabs}

@@

\title{
	Time stretching of the GeV emission of GRBs: Fermi LAT data vs geometrical model
}

\author{
	Maxim S. Piskunov$^{1,2}$\thanks{\href{mailto:maxit@ms2.inr.ac.ru}{maxit@ms2.inr.ac.ru}} ~and
	Grigory I. Rubtsov$^{2,3}$\thanks{\href{mailto:grisha@ms2.inr.ac.ru}{grisha@ms2.inr.ac.ru}} \\
	{\small $^{1}$Faculty of Physics, Moscow State University, Moscow, Russia}\\
	{\small $^{2}$Institute for Nuclear Research of the RAS, Moscow, Russia}\\
	{\small $^{3}$Novosibirsk State University, Novosibirsk, Russia}
}

@@

\begin{abstract}
	It is known that the high energy $(> \unit[100]{MeV})$
        emission of gamma-ray bursts is delayed with respect to the
        low energy emission.  However, the dependence of light curves
        on energy has not been studied for the high energy bands.  In
        this paper we consider the bursts observed by Fermi-LAT from
        2008 August 4 to 2011 August 1, for which at least $10$
        photons were observed with energy greater than
        $\unit[1]{GeV}$. These include $4$ bursts: GRB 080916C, GRB
        090510, GRB 090902B, and GRB 090926A.  We use the
        Kolmogorov-Smirnov test to compare the light curves in the two
        bands, $\unit[100]{MeV} < E < \unit[1]{GeV}$ and
        $\unit[1]{GeV} < E < \unit[300]{GeV}$. For GRB 080916C and GRB
        090510 the light curves in two bands are statistically
        compatible. However, for GRB 090926A, the higher-energy light
        curve is stretched compared to the lower-energy one with a
        statistical significance of $3.3 \sigma$ and, for GRB 090902B,
        on the contrary, the lower-energy curve is stretched with $2.3
        \sigma$ significance. We argue that the observed diversity of
        stretching factors may be explained in a simple geometrical
        model. The model assumes that the jet opening angle depends on
        the emission energy in a way that the most energetic photons
        are radiated near the axis of the jet. All the bursts are
        considered equivalent in their rest frames and the observed
        light curves differ only due to different redshifts and view
        directions. The model conforms to the total burst energy
        constraint and matches the Fermi LAT observations of the
        fraction of GRBs visible in $\unit[100]{MeV} < E <
        \unit[1]{GeV}$ band, which may be observed at higher
        energies. The model predicts the distribution of observable
        stretching factors, which may be tested in the future
        data. Finally, we propose a way to estimate observer's
        off-axis angle based on stretching factor and fraction of the
        high-energy photons. The code for modeling is open source and
        is publicly available on GitHub
        (\url{https://github.com/maxitg/GammaRays}).
\end{abstract}

%TODO: add grants
%? ref to Lemoine GRB1

@@

Gamma-ray bursts (GRBs) are among the most energetic events in the
Universe. Therefore, the studies of the emission mechanism and
phenomenology of GRBs may provide the new knowledge in particle
physics. Moreover, the GRBs are observed from cosmological distances
and therefore bear an imprint of the late-time evolution of the
Universe. An extensive studies of these explosions led to a number of
interesting results, see~\cite{Vianello:2013ela,Gehrels:2013xd} for a
review. In particular, the total energy emitted in gamma-rays during a
burst was found to be similar for the different GRBs within an order
of magnitude~\cite{Bloom:2003wy}. This indicates that the most of the
bursts have similar energetics in their rest frames. Several
observations are related to the temporal variations of spectra. This
way, the spectral lags were found between different low energy bands
\cite{Yi:2005ht} and the very-high-energy radiation was discovered to
be extended relative to the x-ray emission
\cite{Castignani:2014gaa,Lange:2013uh,Vianello:2013ela}.

%% change the next statement
	We decided to elaborate on that result, and, using the data
        from Large Area Telescope (LAT) of Fermi satellite~\cite{Ackermann:2012kna}, explore the spectral variations between different high energy bands, specifically $(\unit[100]{MeV}, \unit[1]{GeV})$ and $(\unit[1]{GeV}, \unit[300]{GeV})$ (we call them low and high energy bands throughout the paper).
	In particular, we use the Kolmogorov-Smirnov test to compute the time stretching of radiation in one of these bands compared to the other.

@@

\bibliographystyle{ieeetr}
\bibliography{gammaRays.bib}
